The Proofs

### 2023-11-13 proofs/tex/subst-algo-lemmas.tex
Restructure proofs

\begin{lemma}[Algorithmic Substitution Strengthening]
  \label{lemma:subst-restr-uv}
  Restricting the substitution to the algorithmic 
  variables of the
  substitution subject does not affect the result.
  Suppose that $[[uσ]]$ is an algorithmic substitution,
  $[[uP]]$ and $[[uN]]$ are algorithmic types. Then
 \begin{itemize}
    \item[$+$] $[[ [uσ]uP ]] = [[ [uσ|uv uP]uP ]]$,
    \item[$-$] $[[ [uσ]uN ]] = [[ [uσ|uv uN]uN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
\end{proof}

\begin{lemma}[Substitutions equal on the algorithmic variables]
  \label{lemma:subst-eq-algovar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are 
  normalized substitutions of signature $[[Θ ⊢ uσi : Ξ]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a normalized type $[[Γ; Ξ ⊢ uP]]$, if $[[ [uσ1]uP = [uσ2]uP ]]$ then
      $[[uσ1|(uv uP)  = uσ2|(uv uP)]]$;
    \item [$-$] for a normalized type $[[Γ; Ξ ⊢ uN]]$, if $[[ [uσ1]uN = [uσ2]uN ]]$ then
      $[[uσ1|(uv uN)  = uσ2|(uv uN)]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  The proof is a simple structural induction on 
  $[[Γ; Ξ ⊢ uP]]$ and mutually, on $[[Γ; Ξ ⊢ uN]]$.
  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
  \begin{caseof}
    \item $[[Γ; Ξ ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
    \item $[[Γ; Ξ ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
    \item $[[Γ; Ξ ⊢ ∀pas.uN]]$. Then we are proving that
      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; Ξ ⊢ uN]]$, 
      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
      as required.
    \item $[[Γ; Ξ ⊢ uP → uN]]$. Then we are proving that
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
      By definition of substitution and congruence of equality, 
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
      and well-formed in the same contexts.
      This way, by the induction hypothesis, 
      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
    \item $[[Γ; Ξ ⊢ ↑uP]]$. The proof is similar to the previous case:
      we apply congruence of substitution, equality, and normalization,
      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
  \end{caseof}
\end{proof}

\begin{corollary}[Substitutions equivalent on the algorithmic variables]
  \label{lemma:subst-equiv-algovar}
  Suppose that $[[uσ1]]$ and $[[uσ2]]$ are substitutions 
  of signature $[[Θ ⊢ uσi : Ξ]]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a type $[[Γ; Ξ ⊢ uP]]$, if $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ then
      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uP]]$;
    \item [$-$] for a type $[[Γ; Ξ ⊢ uN]]$, if $[[Θ ⊢ [uσ1]uN ≈ [uσ2]uN]]$ then
      $[[Θ ⊢ uσ1 ≈ uσ2 : uv uN]]$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  First, let us normalize the types and the substitutions, and show that
  the given equivalences and well-formedness properties are preserved. 
  $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
  \cref{lemma:subt-equiv-algorithmization}.
  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
  Notice that by \cref{corollary:norm-subst-sig-algo}
  $[[Θ ⊢ uσi : Ξ]]$ implies $[[Θ ⊢ nf(uσi) : Ξ]]$.

>>>>

\lemmaSubstRestrUV*
\begin{proof}
  The proof is analogous to the proof of \cref{lemma:subst-restr-fv}.
\end{proof}

\lemmaSubstEqAlgVar*
\begin{proof}
  The proof is a simple structural induction on 
  $[[Γ; Ξ ⊢ uP]]$ and mutually, on $[[Γ; Ξ ⊢ uN]]$.
  Let us consider the shape of $[[uN]]$ (the cases of $[[uP]]$ are symmetric).
  \begin{caseof}
    \item $[[Γ; Ξ ⊢ α̂⁻]]$. Then $[[ [uσ1]α̂⁻ = [uσ2]α̂⁻ ]]$
      implies $[[uσ1|(uv α̂⁻) = uσ2|(uv α̂⁻)]]$ immediately.
    \item $[[Γ; Ξ ⊢ α⁻]]$. Then $[[uv α̂⁻ = ∅]]$, and
      $[[uσ1|(uv α̂⁻)  = uσ2|(uv α̂⁻)]]$ holds vacuously.
    \item $[[Γ; Ξ ⊢ ∀pas.uN]]$. Then we are proving that
      $[[ [uσ1]∀pas.uN = [uσ2]∀pas.uN ]]$ implies $[[uσ1|(uv ∀pas.uN) = uσ2|(uv ∀pas.uN)]]$.
      By definition of substitution and \ruleref{\ottdruleEOneForallLabel}, 
      $[[ [uσ1]uN = [uσ2]uN ]]$ implies $[[uσ1|uv uN = uσ2|uv uN]]$.
      Since $[[∀pas.uN]]$ is normalized, so is $[[Γ, pas; Ξ ⊢ uN]]$, 
      hence, the induction hypothesis is applicable and implies $[[uσ1|uv uN = uσ2|uv uN]]$,
      as required.
    \item $[[Γ; Ξ ⊢ uP → uN]]$. Then we are proving that
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$ implies $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
      By definition of substitution and congruence of equality, 
      $[[ [uσ1](uP → uN) = [uσ2](uP → uN) ]]$
      means $[[ [uσ1]uP = [uσ2]uP ]]$ and $[[ [uσ1]uN = [uσ2]uN ]]$.
      Notice that $[[uP]]$ and $[[uN]]$ are normalized since $[[uP → uN]]$ is normalized, 
      and well-formed in the same contexts.
      This way, by the induction hypothesis, 
      $[[uσ1|(uv uP) = uσ2|(uv uP)]]$ and $[[uσ1|(uv uN) = uσ2|(uv uN)]]$,
      which since $[[uv (uP → uN) = uv uP ∪ uv uN]]$ implies
      $[[uσ1|(uv uP → uN) = uσ2|(uv uP → uN)]]$.
    \item $[[Γ; Ξ ⊢ ↑uP]]$. The proof is similar to the previous case:
      we apply congruence of substitution, equality, and normalization,
      then the induction hypothesis, and then the fact that $[[uv (↑uP) = uv uP]]$.
  \end{caseof}
\end{proof}

\corollarySubstEquivAlgVar*
\begin{proof}
  First, let us normalize the types and the substitutions, and show that
  the given equivalences and well-formedness properties are preserved. 
  $[[Γ; Ξ ⊢ uP]]$ implies $[[Γ; Ξ ⊢ nf(uP)]]$ by \cref{corollary:wf-nf-algo}.
  $[[Θ ⊢ [uσ1]uP ≈ [uσ2]uP]]$ implies $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ by 
  \cref{lemma:subt-equiv-algorithmization}.
  Then $[[ nf([uσ1]uP) = nf([uσ2]uP)]]$ implies 
  $[[ [nf(uσ1)]nf(uP) = [nf(uσ2)]nf(uP)]]$ by \cref{lemma:norm-subst-distr}.
  Notice that by \cref{corollary:norm-subst-sig-algo}
  $[[Θ ⊢ uσi : Ξ]]$ implies $[[Θ ⊢ nf(uσi) : Ξ]]$.

### 2023-11-13 proofs/tex/subst-lemmas.tex
Restructure
Restructure proofs

\begin{lemma}[Substitution strengthening]
  \label{lemma:subst-restr-fv}
  Restricting the substitution to the free variables of the
  substitution subject does not affect the result.
  Suppose that $[[σ]]$ is a substitution,
  $[[iP]]$ and $[[iN]]$ are types. Then
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  First, we strengthen the statement by saying that
  one can restrict the substitution to an arbitrary
  superset of the free variables of the substitution subject:
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
  \end{itemize}
  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
  For the base cases: 
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
    \item $[[iN = iP → iM]]$\\
      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
      Then 
      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
       = [[ [σ|varset](iP → iM) ]]$.
    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
      Let us assume $[[pas]]$ are fresh variables, 
      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
      $[[α± ]] \in [[pas]]$, and thus, 
      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
      immediately from the definition.

@@

\begin{lemma}[Signature of a restricted substitution]
  \label{lemma:subst-restr-sig}
  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2 ⊢ σ|varset :  Γ1 ∩ varset]]$.
\end{lemma}
\begin{proof}
  Let us take an arbitrary $[[α± ∊ Γ1 ∩ varset]]$.
  Since $[[α± ∊ Γ1]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.

@@

\begin{lemma}
  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. 
  Then if $[[varset]]$ is disjoint from $[[Γ1]]$, 
  then $[[ σ|varset]] = [[ id ]]$. 
\end{lemma}
\begin{proof}
  Let us take an arbitrary $[[α±]]$.
  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α±]] = [[α±]]$ by definition.

@@

\begin{corollary}[Application of a disjoint substitution]
  \label{corollary:subst-disj}
  Suppose that $[[σ]]$ is a substitution with signature $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
    \item [$+$] if $[[Γ1 ∩ fv(iQ) = ∅]]$ then $[[ [σ] iQ = iQ]]$;
    \item [$-$] if $[[Γ1 ∩ fv(iN) = ∅]]$ then $[[ [σ] iN = iN]]$.
  \end{itemize}
\end{corollary}

\begin{lemma}[Substitution range weakening]
  \label{lemma:subst-range-weakening}
  Suppose that $[[Γ2 ⊆ Γ'2]]$ are contexts and 
  $[[σ]]$ is a substitution.
  Then $[[Γ2 ⊢ σ : Γ1]]$ implies $[[Γ'2 ⊢ σ : Γ1]]$.
\end{lemma}
\begin{proof}
  For any $[[α± ∊ Γ1]]$, 
  $[[Γ2 ⊢ σ : Γ1]]$ gives us  $[[ Γ2 ⊢ [σ]α± ]]$, which can be
  weakened to $[[ Γ'2 ⊢ [σ]α± ]]$ by \cref{lemma:wf-weakening}. 
  This way, $[[Γ'2 ⊢ σ : Γ1]]$.
\end{proof}

\begin{lemma}[]
  Suppose that $[[Γ' ⊆ Γ]]$,
  $[[σ1]]$ and $[[σ2]]$ are substitutions of signature $[[Γ ⊢ σi : Γ']]$.
  Then 
  \begin{enumerate}
    \item [$+$] for a type $[[Γ ⊢ iP]]$, if $[[Γ ⊢ [σ1]iP ≈ [σ2]iP]]$ then 
    $[[Γ ⊢ σ1 ≈ σ2 : fv iP ∩ Γ']]$;
    \item [$-$] for a type $[[Γ ⊢ iN]]$, if $[[Γ ⊢ [σ1]iN ≈ [σ2]iN]]$ then
    $[[Γ ⊢ σ1 ≈ σ2 : fv iN ∩ Γ']]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Let us make an additional assumption that $[[σ1]]$, $[[σ2]]$, 
  and the mentioned types are normalized. If they are not,
  we normalize them first.

@@

\begin{lemma}[Substitution composition well-formedness]
  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
  then $[[Γ'1, Γ'2 ⊢ σ2 ○ σ1 : Γ1, Γ2]]$.
\end{lemma}

\begin{lemma}[Substitution monadic composition well-formedness]
  \label{lemma:subst-monad-composition-wf}
  If $[[Γ'1 ⊢ σ1 : Γ1]]$ and $[[Γ'2 ⊢ σ2 : Γ2]]$,
  then $[[Γ'2 ⊢ σ2 <=< σ1 : Γ1]]$.
\end{lemma}

\begin{lemma}[Substitution composition]
  \label{lemma:subst-composition}
    If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, 
    $[[Γ1 ∩ Γ'2 = ∅ ]]$ and $[[ Γ1 ∩ Γ2 = ∅ ]]$ then 
    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
\end{lemma}

\begin{corollary}[Substitution composition commutativity]
  \label{corollary:subst-composition-commutativity}
  If $[[Γ'1 ⊢ σ1 : Γ1]]$, $[[Γ'2 ⊢ σ2 : Γ2]]$, and
  $[[ Γ1 ∩ Γ2 = ∅ ]]$, 
  $[[ Γ1 ∩ Γ'2 = ∅ ]]$, and
  $[[ Γ'1 ∩ Γ2 = ∅ ]]$ then 
  $[[ σ2 ○ σ1 ]] = [[ σ1 ○ σ2 ]]$.
\end{corollary}
\begin{proof}
  by \cref{lemma:subst-composition},
    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
    Since the codomain of $[[σ1]]$ is $[[Γ'1]]$,
    and it is disjoint with the domain of $[[σ2]]$,
    $[[σ2 <=< σ1]] = [[σ1]]$.
\end{proof}

\begin{lemma}[Substitution domain weakening]
  \label{lemma:subst-domain-weakening}
  If $[[Γ2 ⊢ σ : Γ1]]$ then $[[Γ2, Γ' ⊢ σ : Γ1, Γ']]$
\end{lemma}
\begin{proof}
  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption,
  and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ']] ⊆ [[Γ2, Γ']]$, 
  and thus, $[[Γ2, Γ' ⊢ α± ]]$.
\end{proof}

\begin{lemma}[Free variables after substitution]
  \label{lemma:subst-fv}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, then
  \begin{enumerate}
    \item [$+$] for a type $[[iP]]$,
      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
     $[[fv(iP) \ Γ1 ⊆ fv([σ]iP) ⊆ (fv(iP) \ Γ1) ∪ Γ2]]$;
    \item [$-$] for a type $[[iN]]$, 
      the free variables of $[[ [σ]iP ]]$ are bounded in the following way:
      $[[fv(iN) \ Γ1 ⊆ fv([σ]iN) ⊆ (fv(iN) \ Γ1) ∪ Γ2]]$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We prove it by structural induction on 
  $[[iP]]$ and mutually, on $[[iN]]$.
  \begin{caseof}
    \item $[[iP = α⁺]]$\\
      If $[[α⁺ ∊ Γ1]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ Γ2]]$.
      $[[fv(α⁺) \ Γ1]] = [[∅]]$, so $[[fv([σ]iP) \ Γ1  ⊆ fv([σ]α⁺)]]$
      vacuously.

@@

\begin{lemma}[Free variables of a variable image]
  \label{lemma:subst-fv-image}
  Suppose that $[[σ]]$ is an arbitrary substitution, 
  Then 
  \begin{itemize}
    \item [$+$] if $[[α± ∊ fv(iP)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iP) ]]$,
    \item [$-$] if $[[α± ∊ fv(iN)]]$ then $[[ fv([σ]α±) ⊆ fv([σ]iN) ]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By mutual induction on $[[iP]]$ and $[[iN]]$.
  The base cases (when $[[iP]]$ or $[[iN]]$ is a variable) 
  are trivial, since then $[[α± ∊ fv(iP)]]$ means 
  $[[α± = iP]]$ (and symmetrically for $[[iN]]$).
  The congruent cases 
  (when the type is formed by $[[↓]]$, $[[↑]]$, or $[[→]]$)
  hold since $[[α±]]$ occurs in type means that it occurs
  in one of its parts, to which we apply the induction hypothesis.

  Let us suppose that te type is $[[∃nas.iQ]]$.
  Then $[[α± ∊ fv(∃nas.iQ)]]$ means
  $[[α± ∊ fv(iQ)]]$ and $[[α± ∉ {nas}]]$.
  Then by the induction hypothesis, 
  $[[ fv([σ]α±) ⊆ fv([σ]iQ) ]]$,
  and it is left to notice that $[[ fv([σ]α±) ∩ {nas} = ∅ ]]$,
  which we can ensure by alpha-equivalence.
\end{proof}

>>>>

\lemmaSubstRestrFv*
\begin{proof}
  First, we strengthen the statement by saying that
  one can restrict the substitution to an arbitrary
  superset of the free variables of the substitution subject:
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
  \end{itemize}
  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
  For the base cases: 
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
    \item $[[iN = iP → iM]]$\\
      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
      Then 
      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
       = [[ [σ|varset](iP → iM) ]]$.
    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
      Let us assume $[[pas]]$ are fresh variables, 
      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
      $[[α± ]] \in [[pas]]$, and thus, 
      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
      immediately from the definition.

@@

\lemmaSubstRestrSig*
\begin{proof}
  Let us take an arbitrary $[[α± ∊ Γ1 ∩ varset]]$.
  Since $[[α± ∊ Γ1]]$, $[[Γ2 ⊢ [σ]α± ]]$ by the signature of $[[σ]]$.

@@

\lemmaSubstDisjoint*
\begin{proof}
  Let us take an arbitrary $[[α±]]$.
  If $[[α± ∉ varset]]$ then $[[ [σ|varset]α±]] = [[α±]]$ by definition.

@@

\corollarySubstDisj*

\lemmaSubstRangeWeak*
\begin{proof}
  For any $[[α± ∊ Γ1]]$, 
  $[[Γ2 ⊢ σ : Γ1]]$ gives us  $[[ Γ2 ⊢ [σ]α± ]]$, which can be
  weakened to $[[ Γ'2 ⊢ [σ]α± ]]$ by \cref{lemma:wf-weakening}. 
  This way, $[[Γ'2 ⊢ σ : Γ1]]$.
\end{proof}

\lemmaSubstEquiv*
\begin{proof}
  Let us make an additional assumption that $[[σ1]]$, $[[σ2]]$, 
  and the mentioned types are normalized. If they are not,
  we normalize them first.

@@

\lemmaSubstCompWell*
\begin{proof}
    By definition of composition.
\end{proof}

\lemmaSubstMonadCompWell*
\begin{proof}
  By definition of monadic composition. 
\end{proof}

\lemmaSubstComp*
\begin{proof}
  \hfill
  \begin{enumerate}
    \item Suppose that $[[α± ∊ Γ1]]$ then $[[α± ∉ Γ2]]$ , and thus, 
      $[[ [(σ2 <=< σ1) ○ σ2]α± = [(σ2 <=< σ1)]α± ]] = [[ [σ2][σ1]α± = [(σ2 ○ σ1)]α± ]]$.
    \item Suppose that $[[α± ∉ Γ1]]$ then $[[ [(σ2 ○ σ1)]α± = [σ2]α± ]]$.  Then 
    \begin{enumerate}
      \item if $[[α± ∉ Γ2]]$ then $[[ [σ2]α± = α± ]]$ and 
        $[[ [(σ2 <=< σ1) ○ σ2]α± = [(σ2 <=< σ1)][σ2]α±]] = [[ [σ2 <=< σ1]α± = α± ]]$
      \item if  $[[α± ∊ Γ2]]$ then $[[ Γ2' ⊢ [σ2]α±  ]]$, and hence, 
        $[[ [(σ2 <=< σ1) ○ σ2]α±  = [(σ2 <=< σ1)][σ2]α±]] = [[ [σ2]α± ]]$
        by definition of monadic composition, since 
        none of the free variables of  $[[ [σ2]α± ]]$ is in $[[Γ1]]$.
    \end{enumerate}
  \end{enumerate}
\end{proof}

\corollarySubstCompCumm*
\begin{proof}
  by \cref{lemma:subst-composition},
    $[[ σ2 ○ σ1 ]] = [[ (σ2 <=< σ1) ○ σ2 ]]$.
    Since the codomain of $[[σ1]]$ is $[[Γ'1]]$,
    and it is disjoint with the domain of $[[σ2]]$,
    $[[σ2 <=< σ1]] = [[σ1]]$.
\end{proof}

\lemmaSubstDomWeak*
\begin{proof}
  If the variable $[[α±]]$ is in $[[Γ1]]$ then $[[Γ2 ⊢ [σ]α± ]]$ by assumption,
  and then $[[Γ2, Γ' ⊢ [σ]α± ]]$ by weakening.
  If the variable $[[α±]]$ is in $[[Γ' \ Γ1]]$ then $[[ [σ]α± = α± ]] \in [[Γ' ⊆ Γ2, Γ']]$, 
  and thus, $[[Γ2, Γ' ⊢ α± ]]$.
\end{proof}

\lemmaSubstFv*
\begin{proof}
  We prove it by structural induction on 
  $[[iP]]$ and mutually, on $[[iN]]$.
  \begin{caseof}
    \item $[[iP = α⁺]]$\\
      If $[[α⁺ ∊ Γ1]]$ then $[[Γ2 ⊢ [σ]α⁺]]$, and by 
      \cref{lemma:wf-soundness}, $[[fv([σ]α⁺) ⊆ Γ2]]$.
      $[[fv(α⁺) \ Γ1]] = [[∅]]$, so $[[fv([σ]iP) \ Γ1  ⊆ fv([σ]α⁺)]]$
      vacuously.

@@

\lemmaSubstFvImage*
\begin{proof}
  By mutual induction on $[[iP]]$ and $[[iN]]$.
  The base cases (when $[[iP]]$ or $[[iN]]$ is a variable) 
  are trivial, since then $[[α± ∊ fv(iP)]]$ means 
  $[[α± = iP]]$ (and symmetrically for $[[iN]]$).
  The congruent cases 
  (when the type is formed by $[[↓]]$, $[[↑]]$, or $[[→]]$)
  hold since $[[α±]]$ occurs in type means that it occurs
  in one of its parts, to which we apply the induction hypothesis.

  Let us suppose that the type is $[[∃nas.iQ]]$.
  Then $[[α± ∊ fv(∃nas.iQ)]]$ means
  $[[α± ∊ fv(iQ)]]$ and $[[α± ∉ {nas}]]$.
  Then by the induction hypothesis, 
  $[[ fv([σ]α±) ⊆ fv([σ]iQ) ]]$,
  and it is left to notice that $[[ fv([σ]α±) ∩ {nas} = ∅ ]]$,
  which we can ensure by alpha-equivalence.
\end{proof}
